### 2022-12-09 dissertation/chapters/03-heuristics.tex
✨ (symbols): Chapter 2 done
✨ (symbols): Chapters 3,4,7 done

where $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function $\mathcal{E}$, relative to the weight at index $i$. The learning rate controls the step-size that is taken in the direction of the negative gradient at each time step. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight $w$, at index $i$, is given as

\begin{equation}
      \label{eq:heuristics:gd:update_step_part_3}
      \frac{\partial \epsilon}{\partial w_{i}} = -2(t_{p} - o_{p})\frac{\partial f}{\partial net_{p}}z_{i,p}
\end{equation}

where $t_{p}$ refers to the target value for pattern $p$, $o_{p}$ refers to the output value for pattern $p$, $f$ refers to the \index{activation function}activation function, $net_{p}$ refers to the \index{net input signal}net input signal for pattern $p$, and $z_{i,p}$ refers to the input value at index $i$ for pattern $p$.

@@

A simplified notation is proposed from which the gradient-based heuristics can be compared. Consider a simplification of the update-step for \acs{SGD} as presented in Equations~\eqref{eq:heuristics:gd:update_step_part_1} and~\eqref{eq:heuristics:gd:update_step_part_2} where $\frac{\partial \epsilon}{\partial w_{i}}$ is simply represented as the gradient term $g$, resulting in a simplified weight update step and is given as

@@

\begin{equation}
      \label{eq:heuristics:gd:adagrad_part_2}
      \begin{split}
            w_{t+1} = w_{t} - \frac{\eta}{\sqrt{G_{t,i} + \epsilon}} \odot g_{t,i}
      \end{split}
\end{equation}

@@

In Equations~\eqref{eq:heuristics:gd:adam_part_1} and~\eqref{eq:heuristics:gd:adam_part_2} above, $\beta_{1}$ and $\beta_{2}$ are decay rates, similar to $\alpha$ for \acs{Momentum}. \citeauthor{ref:kingma:2014}~\cite{ref:kingma:2014} suggest default values $B_{1}=0.9$, $B_{2}=0.999$ and $\epsilon = 1 \times 10^{-8}$.

@@

During initialisation, particles are randomly placed within the search space by sampling from a uniform distribution such that $\boldsymbol{x}_{i} \sim U(x_{min}, x_{max})$ and the velocity is set to 0. At time step 0, the particle's initial position is set to be the particle's personal best solution such that $\boldsymbol{y}_{i}(0) = \boldsymbol{x}_{i}(0)$. The particle's update step is then broken into two parts, including a velocity update step presented in Equation~\eqref{eq:heuristics:pso:velocity} followed by a position update step as presented in Equation~\eqref{eq:heuristics:pso:position}.

@@

\citeauthor{ref:vanwyk:2014}~\cite{ref:vanwyk:2014} mentions that when \acp{PSO} where first developed, it was possible for particle velocities to become inappropriately large during optimisation, leading to situations where particles fly out of the feasible search space. This is known as \textit{swarm explosion} and occurs when there are frequent changes in the global best position. In order to address this issues, the concept of \textit{velocity clamping} was introduced~\cite{ref:eberhart:1996}. The idea behind velocity clamping is to restrict particle velocities to some $V_{max}$ threshold, modelling a form of terminal velocity. Velocity clamping is applied after the velocity update step and is given in Equation~\eqref{eq:heuristics:pso:velocity_clamping} below.

@@

\begin{algorithm}[htb]
      \caption{The pseudo-code algorithm for the gbest \acs{PSO} heuristic.}
      \label{algo:heuristics:pso:gbest}
      \begin{algorithmic}
            \State Create swarm of $N$ entities, each with $J$ dimensions;
            \While{stopping condition are not met}
            \For{each particle $i = 1, \dots, N$}
            \State // Set the personal best position;
            \If{$f(x_{i}) < f(y_{i})$ }
            \State $y_{i} = x_{i}$;
            \EndIf

            \State // Set the global best position;
            \If{$f(x_{i}) < f(\hat{y})$ }
            \State $\hat{y} = x_{i}$;
            \EndIf
            \EndFor

@@

\begin{equation}
      \label{eq:heuristics:de:mutation}
      \begin{split}
            \boldsymbol{u}_{i}(t) = \boldsymbol{x}_{i_{1}} + \beta(\boldsymbol{x}_{i_{2}}(t) - \boldsymbol{x}_{i_{3}}(t))
      \end{split}
\end{equation}

>>>>

where $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function $\mathcal{E}$, relative to the weight at index $i$. The learning rate controls the step-size that is taken in the direction of the negative gradient at each time step. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight vector $\boldsymbol{w}$, at index $i$, is given as

\begin{equation}
	\label{eq:heuristics:gd:update_step_part_3}
	\frac{\partial \epsilon}{\partial w_{i}} = -2(t_{i,p} - o_{i,p})\frac{\partial f}{\partial net_{p}}z_{i,p}
\end{equation}

where $t_{i,p}$ refers to the $i$-th element of the target value for pattern $p$, $o_{i,p}$ refers to the $i$-th element of the output value for pattern $p$, $f$ refers to the \index{activation function}activation function, $net_{p}$ refers to the \index{net input signal}net input signal for pattern $p$, and $z_{i,p}$ refers to the input value at index $i$ for pattern $p$.

@@

A simplified notation is proposed from which the gradient-based heuristics can be compared. Consider a simplification of the update-step for \acs{SGD} as presented in Equations~\eqref{eq:heuristics:gd:update_step_part_1} and~\eqref{eq:heuristics:gd:update_step_part_2} where $\frac{\partial \epsilon}{\partial w_{i}}$ is simply represented as the gradient term $\boldsymbol{g}$, resulting in a simplified weight update step and is given as

@@

\begin{equation}
	\label{eq:heuristics:gd:adagrad_part_2}
	\begin{split}
		w_{t+1,i} = w_{t,i} - \frac{\eta}{\sqrt{G_{t,i} + \epsilon}} \odot g_{t,i}
	\end{split}
\end{equation}

@@

In Equations~\eqref{eq:heuristics:gd:adam_part_1} and~\eqref{eq:heuristics:gd:adam_part_2} above, $\beta_{1}$ and $\beta_{2}$ are decay rates, similar to $\alpha$ for \acs{Momentum}. \citeauthor{ref:kingma:2014}~\cite{ref:kingma:2014} suggest default values $\beta_{1}=0.9$, $\beta_{2}=0.999$ and $\epsilon = 1 \times 10^{-8}$.

@@

During initialisation, particles are randomly placed within the search space by sampling from a uniform distribution such that $\boldsymbol{x}_{i} \sim U(\boldsymbol{x}_{min}, \boldsymbol{x}_{max})$ and the velocity is set to 0. At time step 0, the particle's initial position is set to be the particle's personal best solution such that $\boldsymbol{y}_{i}(0) = \boldsymbol{x}_{i}(0)$. The particle's update step is then broken into two parts, including a velocity update step presented in Equation~\eqref{eq:heuristics:pso:velocity} followed by a position update step as presented in Equation~\eqref{eq:heuristics:pso:position}.

@@

\citeauthor{ref:vanwyk:2014}~\cite{ref:vanwyk:2014} mentions that when \acp{PSO} where first developed, it was possible for particle velocities to become inappropriately large during optimisation, leading to situations where particles fly out of the feasible search space. This is known as \textit{swarm explosion} and occurs when there are frequent changes in the global best position. In order to address this issues, the concept of \textit{velocity clamping} was introduced~\cite{ref:eberhart:1996}. The idea behind velocity clamping is to restrict particle velocities to some $\boldsymbol{V_{max}{}$ threshold, modelling a form of terminal velocity. Velocity clamping is applied after the velocity update step and is given in Equation~\eqref{eq:heuristics:pso:velocity_clamping} below.

@@

\begin{algorithm}[htb]
	\caption{The pseudo-code algorithm for the gbest \acs{PSO} heuristic.}
	\label{algo:heuristics:pso:gbest}
	\begin{algorithmic}
		\State Create swarm of $N$ entities, each with $J$ dimensions;
		\While{stopping condition are not met}
		\For{each particle $i = 1, \dots, N$}
		\State // Set the personal best position;
		\If{$f(\boldsymbol{x}_{i}) < f(\boldsymbol{y}_{i})$ }
		\State $\boldsymbol{y}_{i} = \boldsymbol{x}_{i}$;
		\EndIf

		\State // Set the global best position;
		\If{$f(\boldsymbol{x}_{i}) < f(\boldsymbol{\hat{y}})$ }
		\State $\boldsymbol{\hat{y}} = \boldsymbol{x}_{i}$;
		\EndIf
		\EndFor

@@

\begin{equation}
	\label{eq:heuristics:de:mutation}
	\begin{split}
		\boldsymbol{u}_{i}(t) = \boldsymbol{x}_{i_{1}}(t) + \beta(\boldsymbol{x}_{i_{2}}(t) - \boldsymbol{x}_{i_{3}}(t))
	\end{split}
\end{equation}
